---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

_This post will differ from the rest of the blog in the sense that it will illustrate some Data Vizualization techniques, an aspect of my day to day job as a Data Analyst._

_In a further post, once I will be satisfied enough with the shape and form of my website, I will outline the various steps I undertook to create this site in its current form (spoiler: lots of tweaks of existing templates that were placed together in here)_

## 0. Load libraries
```{r libs,warning=F, message=F, comment=F, results=F}
library(tidyverse)
library(lubridate)
```

## 1. Load the data
The first thing we do is load the data. If you're not familiar with R, be aware that you can directly load the data from a Github repository by linking it to its raw page as we do here:
```{r data,warning=FALSE, message=FALSE}
trains_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/full_trains.csv")

small_trains <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/small_trains.csv") 
```


## 2. Analysis

We now have access to our data and are eager to start exploring it. However, we must first reflect upon the range of possibilities that emanate from such a dataset. 

### Exploring the causes of delays for the top 10 train stations in France.

In order to avoid getting confused with the amount of angles we could dive into, it can be helpful to start with a rather concrete idea of what can be done. There is an infinite set of delay reasons that might be more or less relevants based on the train stations sizes. I wonder if, for the biggest ones, and for the smallest ones, the reason given is similar.

I will therefore attempt the following steps:

1. Isolating the biggest and smallest train stations in terms of passengers
2. Reshaping the delay data 
3. Results in table form
4. Visualization

#### 1. Isolating the biggest train stations in terms of passengers
Let's first see which are the top and lowest stations in terms of total trips. In general, I would avoid some tiny stations with very few passengers by using a threshold of at least 100 trips. But a quick look at the departure stations indicated that there were only 59 stations in the dataset, so a filter has already been applied by the curators of the dataset. We will select the top 10 and bottom 10 as reference groups.

```{r top_stations,warning=F,message=F}
big_stations <- trains_raw %>% 
  group_by(departure_station) %>% 
  summarize(trips=sum(total_num_trips)) %>% 
  arrange(desc(trips)) %>% 
  top_n(10) 

small_stations <- trains_raw %>% 
  group_by(departure_station) %>% 
  summarize(trips=sum(total_num_trips)) %>% 
  arrange(-desc(trips)) %>% 
  top_n(10) 
```

#### 2. Reshaping the delay data 
Now that we know which trainstations we want to look at, we can also:

* Add a dummy that indicates whether they're part of those largest or smallest stations
* Reorganize the "delay reasons" variables
  * Currently, the dataset is in wide format (some variables are spread out) and we want them in long format. More info about this very important nuance can be found [here](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/)

The data ranges from 2015 to 2017. 
Since the delay ratios vary per year, it's important to treat them accordingly.

There are trains going for long trips and there are trains going for short trips. Do they differ? Could this be a better indicator of the delay factors than the simple departure station?


```{r gather_data,warning=F,message=F}
connections <- trains_raw %>% 
  mutate(station_size = case_when(departure_station %in% big_stations$departure_station ~ "BIG",
                                  departure_station %in% small_stations$departure_station ~ "SMALL"),
         connection = str_c(departure_station,arrival_station),
         date_full = str_c(year,month)) %>% 
  filter(is.na(station_size) == F & year == 2017) %>% 
  select(date_full,departure_station,arrival_station,connection,journey_time_avg,total_num_trips,
         num_of_canceled_trains,num_late_at_departure,avg_delay_all_departing,num_arriving_late,
         avg_delay_late_on_arrival,delay_cause_external_cause:delay_cause_travelers) %>% 
  gather(key="delay_cause",value="ratio",starts_with("delay_cause"))

connections %>% 
  arrange(desc(journey_time_avg)) %>% 
  View()
```


